Determinism, Complexity, and Predictability in Computer Performance

### 2014-02-10 pecomplexity.tex
fixing merge errors

Computers are among the most complex engineered artifacts in current
use.  Modern microprocessor chips contain multiple processing units
and multi-layer memories, for instance, and they use complicated
hardware/software strategies to move data and threads of computation
across those resources.  These features---along with all the others
that go into the design of these chips---make the patterns of their
processor loads and memory accesses highly complex and hard to
predict.  Accurate forecasts of these quantities, if one could
construct them, could be used to improve computer design.  If one
could predict that a particular computational thread would be bogged
down for the next 0.6 seconds waiting for data from main memory, for
instance, one could save power by putting that thread on hold for that
time period (e.g., by migrating it to a processing unit whose clock
speed is scaled back).  Computer performance traces are, however, very
complex.  Even a simple ``microkernel,'' like a three-line loop that
repeatedly initializes a matrix in column-major order, can produce
{\sl chaotic} performance traces \cite{mytkowicz09}, as shown in
Figure~\ref{fig:ipc}, and chaos places fundamental limits on
predictability.
%
 \begin{figure}[htbp]
    \centering
    \includegraphics[width=\textwidth]{figs/colshortts}
    % where an .eps filename suffix will be assumed under latex,
    % and a .pdf suffix will be assumed for pdflatex
    \caption{A small snippet of the instructions per cycle(ipc) of {\tt
        \col}, a three-line C program that repeatedly initializes
      a matrix in column-major order, running on an Intel i7\textsuperscript{\textregistered}-based machine.  Even this
      simple program exhibits chaotic performance dynamics.}
   \label{fig:ipc}
  \end{figure}

@@

As a figure of merit we calculate the Mean Absolute Squared Error (MASE)\cite{MASE} between the true and predicted signals: 
%In order to compare the resulting forecasts we calculate the Mean Absolute Squared Error (MASE)\cite{MASE} between the true and predicted signals:
$$MASE = \sum_{j=n+1}^{k+n+1}\frac{|c_j-p_j| }{\frac{k}{n-1}\sum^n_{i=2}|X_{i,obs}-X_{i-1,obs}|}$$
The scaling term for MASE:
$$\frac{1}{n-1}\sum^n_{i=2}|X_{i,obs}-X_{i-1,obs}|$$ 
is the average in-sample forecast error for a random walk prediction $(p_i=X_{i-1,obs})$. This error method was introduced in \cite{MASE} as a ``generally applicable measurement of forecast accuracy without the problems seen in the other measurements." The major advantage of MASE is that it allows fair comparison across methods, prediction horizons and varying signal scales. When a forecast results in a $MASE<1$ this means that the prediction method gave, on average, smaller errors than the 1-step errors from the in-sample random walk forecast strategy. Analogously, $MASE>1$ means that the prediction method did worse, on average than the 1-step errors for the in-sample random walk forecast strategy. In Table \ref{tab:error} we provide the distribution [[Joshua: Ryan, Is this the right word? we give mean $\pm$ std. dev but some have very skewed right tails]]  of MASEs for each of the 8 signals and 3 prediction strategies, these are averaged over 15 runs of each type (signal + method). For comparison Table \ref{tab:error} also has the distribution of weighted permutation entropies for work lengths of $l=5$ and $l=6$.

@@

As can be seen in Table \ref{tab:error} the relationship between prediction accuracy and the weighted permutation entropy (WPE) is much as we conjectured: performance traces with high WPE are
indeed harder to predict using the forecasting models described in Section \ref{sec:compModel}. Figure~\ref{fig:pred_vs_ent} demonstrates this primary finding, with
the exception of one outlier whose behavior we will explain. In Figure~\ref{fig:pred_vs_ent} we plot the best prediction (i.e., the lowest MASE over all 3 methods over all runs of that program) for each of the 8 signals.  
\begin{figure}[htbp]
  \centering
  \includegraphics[width=0.75\textwidth]{figs/prediction_vs_entropy}
  \caption{The best MASE among all runs and prediction methods vs weighted
  permutation entropy. For each of these, the word length used is $6$. The
  dashed line is a least-squares linear fit of all the points except for {\tt
  SVD$_1$} which we have excluded for reasons explained in the text.}
  \label{fig:pred_vs_ent}
\end{figure}
We find that the relationship between the two is roughly linear. The single outlier, \svd regime 1, does not fit the trend due to a weakness in the WPE: namely
that in the absence of any large features, as all the other regimes have, the
WPE effectively falls back to the standard PE for which the noisy behavior
drives toward 1.0.

@@

  %\end{subfigure}
\end{figure} 
In Figures~\ref{fig:lma_pred_vs_ent} and \ref{fig:arima_pred_vs_ent} we directly
compare the performance of the LMA and ARIMA prediction methods (respectively)
to the value of the weighted permutation entropy for all runs of each program
under consideration. The LMA MASE values are largely similar to those of the
best predictions, primarily because LMA often performed superior to ARIMA (and
the na\"ive method). So it is not surprising that the MASE from LMA and the corresponding MASE are highly correlated. On the other hand, the ARIMA MASE values are largely
uncorrelated with WPE values.

The fact that ARIMA is uncorrelated with WPE is in large part one of the major findings of this work. More specifically, say we tried to predict an arbitrary noisy real-valued time series with an ``out-of-the-box" prediction strategy like ARIMA as is proposed as a good strategy in \cite{autoArima} and say we got inconsistent and bad forecasts, (i.e., perform worse than the na\"ive random walk strategy ($MASE>1$). How can we know if the prediction strategy is not adequate for the prediction task or if the signal is simply too complex and unstructured to predict. If a signal is too complex and too little forward information transfer is present we may not be able to do better than the random walk, in which case we should not kill ourselves over finding a more complicated prediction strategy. However, if we measure the complexity to be low, $WPE<0.85$ we can most likely do much better than the random walk and should search for more adequate prediction strategies. For example, consider \col (the blue +s in Figure \ref{fig:lma_vs_arima}). If we use the out-of-the-box ARIMA from \cite{autoArima} we get MASE ranging from 0.5 all the way to 1.0. That is we may think that the best we can do is twice as good as random-walk but sometimes we only do as well as random walk. In which case the random walk prediction strategy is probably your best bet as it is so simple and produces very similar error. However, if we calculate WPE we see that \col has a WPE of $0.5131\pm 0.0034$. This incredibly low entropy implies a great deal of deterministic structure that we can utilize for prediction, even though this did not appear to be the case with ARIMA. In fact using LMA on this signal we get an average MASE of $0.05\pm0.0018$, an error that is on-average 20 times better than the random walk forecast. Alternatively, we can look at the other end of the spectrum at \gcc. This signal has a WPE of $0.9430\pm0.0013$ and each prediction method we applied performed significantly worse than a simple random walk forecast (the best strategy in the case of a structure-less time series. Since \gcc has such little structure according to WPE it is safe to assume that he best strategy will be random walk and that continuing to search for a more accurate prediction strategy will most likely be a fruitless process. Many of the other signals made this jump across the $MASE=1$ line for signals with complexity less than $WPE$ of 0.85. Signals higher than this did not seem to make the jump.

@@

>>>>

Computers are among the most complex engineered artifacts in current
use.  Modern microprocessor chips contain multiple processing units
and multi-layer memories, for instance, and they use complicated
hardware/software strategies to move data and threads of computation
across those resources.  These features---along with all the others
that go into the design of these chips---make the patterns of their
processor loads and memory accesses highly complex and hard to
predict.  Accurate forecasts of these quantities, if one could
construct them, could be used to improve computer design.  If one
could predict that a particular computational thread would be bogged
down for the next 0.6 seconds waiting for data from main memory, for
instance, one could save power by putting that thread on hold for that
time period (e.g., by migrating it to a processing unit whose clock
speed is scaled back).  Computer performance traces are, however, very
complex.  Even a simple ``microkernel,'' like a three-line loop that
repeatedly initializes a matrix in column-major order, can produce
{\sl chaotic} performance traces \cite{mytkowicz09}, as shown in
Figure~\ref{fig:col-ts}, and chaos places fundamental limits on
predictability.
%
 \begin{figure}[htbp]
    \centering
    \includegraphics[width=\textwidth]{figs/colshortts}
    % where an .eps filename suffix will be assumed under latex,
    % and a .pdf suffix will be assumed for pdflatex
    \caption{A small snippet of the instructions per cycle(ipc) of {\tt
        \col}, a three-line C program that repeatedly initializes
      a matrix in column-major order, running on an Intel i7\textsuperscript{\textregistered}-based machine.  Even this
      simple program exhibits chaotic performance dynamics.}
   \label{fig:ipc}
  \end{figure}

\begin{figure}[htbp]
  \centering
  \begin{subfigure}[t]{0.475\textwidth}
    \includegraphics[width=0.95\textwidth]{figs/colshortts}
    \caption{A short segment of the instructions executed per CPU clock cycle
    (IPC) during the execution of \col. Each point is the IPC during a 100,000
    cycle period. There is a great deal of periodicity in this time series.}
    \label{fig:col-ts}
  \end{subfigure}%
  \begin{subfigure}[t]{0.475\textwidth}
    \includegraphics[width=0.95\textwidth]{figs/gccfullts}
    \caption{The full IPC time series during the execution of \gcc. This time
    series has very little structure.}
    \label{fig:gcc-ts}
  \end{subfigure}
  \label{fig:sample-ts}
\end{figure}

@@

As a figure of merit we calculate the Mean Absolute Squared Error (MASE)\cite{MASE} between the true and predicted signals: 
%In order to compare the resulting forecasts we calculate the Mean Absolute Squared Error (MASE)\cite{MASE} between the true and predicted signals:
$$MASE = \sum_{j=n+1}^{k+n+1}\frac{|c_j-p_j| }{\frac{k}{n-1}\sum^n_{i=2}|X_{i,obs}-X_{i-1,obs}|}$$
The scaling term for MASE:
$$\frac{1}{n-1}\sum^n_{i=2}|X_{i,obs}-X_{i-1,obs}|$$ 
is the average in-sample forecast error for a random walk prediction $(p_i=X_{i-1,obs})$. This error method was introduced in \cite{MASE} as a ``generally applicable measurement of forecast accuracy without the problems seen in the other measurements." The major advantage of MASE is that it allows fair comparison across methods, prediction horizons and varying signal scales. When a forecast results in a $MASE<1$ this means that the prediction method gave, on average, smaller errors than the 1-step errors from the in-sample random walk forecast strategy. Analogously, $MASE>1$ means that the prediction method did worse, on average than the 1-step errors for the in-sample random walk forecast strategy. In Table \ref{tab:error} we provide the distribution [[Joshua: Ryan, Is this the right word? we give mean $\pm$ std. dev but some have very skewed right tails]]  of MASEs for each of the 8 signals and 3 prediction strategies, these are averaged over 15 runs of each type (signal + method). For comparison Table \ref{tab:error} also has the distribution of weighted permutation entropies for word lengths of $l=5$ and $l=6$.

@@

As can be seen in Table \ref{tab:error} the relationship between prediction accuracy and the weighted permutation entropy (WPE) is much as we conjectured: performance traces with high WPE are
indeed harder to predict using the forecasting models described in Section \ref{sec:compModel}. Figure~\ref{fig:pred_vs_ent} demonstrates this primary finding, with
the exception of one outlier whose behavior we will explain. In Figure~\ref{fig:pred_vs_ent} we plot the best prediction (i.e., the lowest MASE over all 3 methods over all runs of that program) for each of the 8 signals.

\begin{figure}[htbp]
  \centering
  \includegraphics[width=0.75\textwidth]{figs/prediction_vs_entropy}
  \caption{The best MASE among all runs and prediction methods vs weighted
  permutation entropy. For each of these, the word length used is $6$. The
  dashed line is a least-squares linear fit of all the points except for {\tt
  SVD$_1$} which we have excluded for reasons explained in the text.}
  \label{fig:pred_vs_wpe}
\end{figure}
We find that the relationship between the two is roughly linear. The single
outlier, \svd regime 1, does not fit the trend due to a weakness in the WPE:
namely that in the absence of any large features, as all the other regimes have,
the WPE effectively falls back to the standard PE which the noisy behavior
drives toward 1.0.

@@

  %\end{subfigure}
\end{figure} 
In Figures~\ref{fig:lma_pred_vs_ent} and \ref{fig:arima_pred_vs_ent} we directly
compare the performance of the LMA and ARIMA prediction methods (respectively)
to the value of the weighted permutation entropy for all runs of each program
under consideration. The LMA MASE values are largely similar to those of the
best predictions, primarily because LMA often performed superior to ARIMA (and
the na\"ive method). On the other hand, the ARIMA MASE values are largely
uncorrelated with WPE values. The fact that ARIMA is uncorrelated with WPE is in large part one of the major findings of this work. More specifically, say we tried to predict an arbitrary noisy real-valued time series with an ``out-of-the-box" prediction strategy like ARIMA as proposed in \cite{autoArima} and say we got inconsistent and bad forecasts, (i.e., perform worse than the na\"ive random walk strategy (MASE$>1$). How do we determine if the prediction
strategy is not adequate for the prediction task, or if the signal is simply too
complex to predict. If a signal is too complex and too little forward information transfer is present we may not be able to do better than the random walk, in which case we should not worry ourselves over finding a more complicated prediction strategy. However, if we measure the complexity to be low, $\textrm{WPE}<0.85$ (see Fig.~\ref{fig:pred_vs_wpe}) we can most likely do much better than the random walk and should search for more adequate prediction strategies.

By way of example, consider \col (the blue +s in
Figure~\ref{fig:predictions_vs_wpe}). If we use the out-of-the-box ARIMA from \cite{autoArima} we get MASE ranging from $\sim0.5$ to
$\sim1.0$.  From that information we may assume that the best predictions that
can be done are twice as good as random-walk, and in some cases predictions will
only do as well as random-walk. In the latter such cases, the random-walk
prediction strategy is probably the best bet as it is so simple and produces
very similar error. However, if we calculate the WPE we see that \col has a WPE
of $0.5131 \pm 0.0034$. This incredibly low entropy implies a great deal of
deterministic structure which can be utilize for prediction, even though this
did not appear to be the case with ARIMA. In fact using LMA on this signal we
get an average MASE value of $0.05\pm0.0018$, an error that is on average 20
times better than the random walk forecast, and 10 times better than the best
ARIMA forcast, with little variance. Alternatively, we can look at the other end
of the spectrum: \gcc. This signal has a WPE of $0.9430\pm0.0013$ and each
prediction method we applied was significantly out performed by a simple random
walk (the best strategy in the case of structureless time series). Since \gcc
has such little structure according to WPE it is safe to assume that he best
strategy will be random walk and that continuing to search for a more accurate
prediction strategy will likely be a fruitless process.

@@

%%Now switch form the existence of structure as signaled by WPE to the type of structure

[[Joshua: Merge the following 2 paragraphs.]]

The WPE is sensitive to both linear and nonlinear structure. When you have a low
WPE and a high ARIMA it could be that the structure WPE is picking up is simply
nonlinear structure that LMA can handle but ARIMA cannot. So while the ARIMA
prediction look really bad there is plenty of structure present as suggested by
WPE and taken advantage of by LMA but since it is nonlinear ARIMA can't take it
into account and does bad.

Other noteworthy features of the LMA and ARIMA results are the cluster locations
and distributions. The WPE values of each run any particular program tend to
have little variance, leading to the clusters in
Figures~\ref{fig:arima_pred_vs_wpe}, \ref{fig:lma_pred_vs_wpe} to be fairly
constrained in the $y$ direction. For most traces, the LMA and ARIMA variance is
low too, resulting is small, tight clusters. The ARIMA MASE values of the \col
traces, however, have a large variance resulting in the spread seen in
Fig~\ref{fig:arima_pred_vs_wpe}. Not only are the MASE values of that cluster
bad, in that other predictors vastly outperform it, but they are inconsistent.
Furthermore, since LMA can predict nonlinear behavior while ARIMA can not, we
see that the clusters in Fig.~\ref{fig:arima_pred_vs_wpe} are mostly further to
the right than those in Fig.~\ref{fig:lma_pred_vs_wpe}.
